Category-based and location-based volitional covert attention affect memory at different timescales

### 2023-03-27 paper/main.tex
complete draft-- ready for review!

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\usepackage{setspace,caption}
\doublespacing
\captionsetup[figure]{labelsep=period,labelfont=it,justification=justified,
 singlelinecheck=false,font=doublespacing}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\linenumbers

@@

\author{Kirsten Ziman\textsuperscript{$1, 2$},
Madeline R. Lee$^1$,
Alejandro R. Martinez$^1$,\\
Ethan D. Adner$^1$,
and
Jeremy R. Manning\textsuperscript{$1, \dagger$}\\[0.1in]$^1$Dartmouth College\\
$^2$Princeton University\\
\textsuperscript{$\dagger$}Address correspondence to jeremy.r.manning@dartmouth.edu}

@@

\noindent\textbf{Keywords: covert attention, spatial attention, category-based
attention, recognition memory, perception}

@@

Our brains' cognitive systems detect and exploit patterns in our prior and
ongoing experiences, enabling us to function and adapt in an ever-changing
world. However we do not attend to or treat all types of remembered or incoming
information equally, and our ability to flexibly adapt our thinking and
behaviors can vary markedly with the specific set of concepts or tasks relevant
to a given setting or situation~\citep{ChunTurk07, AlyTurk17, HardNade09,
RangRitc12}. There is also substantial variability across people with respect
to which aspects of experience (sensory, social, emotional, etc.) are noticed,
discriminated between, and acted upon~\citep{HuntEtal89}. This implies that the
same physical (objective) experience may give rise to very different perceived
(subjective) experiences across people~\citep{FreeSimo11, ChanEtal21a}.

The aspects of our experience we attend may be under our volitional control or
may be unconscious or automatic~\citep{JacoEtal92}. Both volitional and
unconscious attention may be expressed overtly, for example through intentional
eye movements~\citep{HoffSubr95} or covertly, without any volitional physical
change~\citep{EngbKlie03}. Prior work has explored the similarities and
differences in the neural basis of overt versus covert
attention~\citep{PosnEtal87, HuntKing03} as well as the behavioral and neural
underpinnings of volitional versus unconscious attention~\citep{DijkAart10} and
their differential effects on memory. There is a general consensus that
sustained volitional attention enhances memory relative to unconscious
attentional processes~\citep{UncaEtal11, TurkEtal13}. However, volitional
attention takes many forms, such as attention to particular spatial locations
or attention to particular visual features or other stimulus properties. How
different \textit{types} of volitional attention combine (or compete) to
enhance memory remains an open question. Volitional covert attention is of
particular interest in that it allows us to dynamically and intentionally
manipulate our experience, even when our sensory input remains largely
static~\citep[i.e., constant physical stimuli, retinal image, etc.][]{YiEtal06,
OcraEtal99}.

Here we examine the ways different types of volitional covert attention
interact to affect memory. We designed an experimental
paradigm~\citep[following][]{Posn80} that asked participants to attend to a
series of presented composite image pairs while keeping their gaze fixed on a
central point. The image pairs comprised a left and right image, each
constructed by blending an image of a face and place. The stimuli and
presentation durations were constant across the two experiments, but the
experiments differed in how often we asked participants to change the focus of
their attention with respect to image category (face versus place) and image
location (left versus right). After the participants attended to a series of
images, we used a recognition memory test to assess which aspects of the
presented images had been encoded into memory. In both experiments we found
that the images participants covertly attended to were better recognized than
other images, supporting the notion that attention enhances memory
encoding~\citep[i.e., they rated attended images as more familiar than
unattended images;][]{Yone02}. After maintaining the focus of attention to a
single image category and location (Sustained Attention Experiment),
participants also recognized the attended-category image at the unattended
location, and (to a lesser extent) the unattended-category image at the
attended location. After more rapidly varying their focus of attention
(Variable Attention Experiment), participants showed a similar boost in
recognition for the unattended-category image at the attended location, but
they did not recognize images at the unattended location. This suggests that
participants were able to shift the location of their covert attentional focus
more rapidly than they were able to shift their focus of covert attention to
stimulus features. We also found differences in the timecourses of these memory
effects, suggesting that the impact of location-based attention on memory
persists on the order of several seconds longer than the impact of
feature-based attention.

@@

We ran a total of 53 participants in across two experimental conditions
(Fig.~\ref{fig:paradigm}). The two conditions differed in how often we
cued participants to change the focus of their attention. All code and
documentation pertaining to our experiments and analyses, along with the
experimental stimuli and data, may be downloaded from
\url{http://www.github.com/ContextLab/attention-memory-task}.

@@

  \caption{\textbf{Experimental paradigm.} \textbf{A.--D. Practice phase.}
  \textbf{A.} Composite face/place image. \textbf{B.} A single pair of
  composite images and a central fixation cross. \textbf{C.} One attention cue
  practice trial. \textbf{D.} Familiarity judgement practice trial. \textbf{E.
  Sustained attention condition.} Participants receive an attention cue,
  followed by a sequence of 10 composite image pairs (Panel E), and then they
  make of familiarity judgements on each of 40 face and place images, presented
  in sequence (Panel F).  \textbf{G.  Variable attention condition.}  Participants
  study a succession of 10 composite image pairs, each preceded by an attention cue.
  After studying the images, they make a series of 40 familiarity judgements about
  face and place images, as in the sustained attention condition (Panel F).}

@@

A total of 53 Dartmouth College undergraduate students enrolled in our study
(30 in the sustained attention condition and 23 in the variable attention
condition). Following a pilot study using a similar experimental design, we
aimed to enroll 30 participants in each condition. However, we fell short of
our enrollment target in the variable attention condition when in-person
testing was discontinued at our institution due to the COVID-19 pandemic. All
participants had self-reported normal or corrected-to-normal vision, memory,
and attention. Participants gave written consent to enroll in the study under a
protocol approved by the Committee for the Protection of Human Subjects at
Dartmouth College.

We used a voluntary pre-experimental survey to collect self-reported
demographic information about each partipant. All 53 participants elected to
fill out the survey. Participants ranged in age from 18--21 years old (mean:
18.7 years; standard deviation: 0.8 years). Participants reported their genders
as female (34 particiapnts), male (18 participants), and gender non-binary (1
participant). Participants reported their ethnicities as Not Hispanic or Latino
(44 particiapnts), Hispanic or Latino (7 participants), or declined to respond
(2 participant). Participants reported their race as White (37 participants),
Asian (13 participants), American Indian or Alaska Native (4 participants),
Black or African American (2 participants), and Other (1 participant). Note
that each participant could report one or more racial categories, as they
deemed appropriate.

Fourty-nine participants reported having no reading impairments, and 4
participants reported having reading impairments such as dyslexia. Fifty
participants reported having normal color vision and 3 reported having abnormal
color vision such as colorblindness. Fifty participants reported taking no
medications and having no recent injuries. One participant reported that they
had recently ``hit [their] head very hard.'' Another reported having taken
concerta in the past, but mentioned they had not taken it recently. One
participant reported using amphetamines regularly, but also clarified that they
were not currently on amphetamines at the time of testing.

We also asked participants to self-report on their sleep, alertness, and coffee
consumption. Participants reported having gotten between 4 and 9 hours of sleep
on the night prior to testing (mean: 6.9 hours; standard deviation: 1.3 hours).
Participants reported their alertness at the time of testing, and we converted
their responses to point values as follows: ``Very alert'' (5 points), ``A
little alert'' (4 points), ``Neutral'' (3 points), ``A little sluggish'' (2
points), and ``Very sluggish'' (1 point). Across all participants, the full
range of alertness values were used (maximum: 5; minimum: 1; mean: 3.44;
standard deviation: 1.0). Participants reported having consumed between 0 and 2
cups of coffee so far on their testing day (mean: 0.3 cups; standard deviation:
0.5 cups).

@@

Participants viewed photographs of faces, places, and composite images each
comprising an equal blend of one face image and one place image. The pool of
360 face images included photographs of adult human male and female faces
selected from the FERET database~\citep{PhilEtal98}. The pool of 360 place
images included photographs of indoor and outdoor places selected from the SUN
database~\citep{XiaoEtal10}. The images we used from both databases came from a
stimulus subset that was manually curated by Megan deBettencourt (personal
communication). All images were resized to $256 \times 256$ pixels, converted
to greyscale, and processed so that every image was matched for mean contrast
and intensity. We selected 20 face images and 20 place images from the stimulus
pool to use in the instructional and practice phases of the experiments
(Fig.~\ref{fig:paradigm}A--D).

@@

 Our experiment was conducted in a sound- and light-attenuated testing room
 containing a chair, desk, and 27-inch iMac desktop computer (display
 resolution: $2048 \times 1152$). The participant sat in the chair and rested
 their chin on a chin rest located 60~cm from the display. The active portion
 of the display screen occupied 52.96$^\circ$ (width) and 31.28$^\circ$
 (height) of the participant's field of view from the chin rest. Stimuli were
 sized to occupy $6.7^\circ$ (width and height) of the participant's field of
 view from the chin rest. We maintained a black background (with any text
 displayed in white) throughout the experiment.

\subsection*{Eyetracking}

We recorded participants' eye gaze positions using a desk-mounted video-based
eyetracker with a spatial resolution of 0.1$^\circ$ visual angle root mean
squared error and a sampling rate of 30~Hz (Eye Tribe, The Eye Tribe,
Copenhagen, Denmark). We calibrated the eyetracker using a 9-point gaze
pattern. As described below, we re-calibrated the eyetracker at regular
intervals throughout the experiment to protect against camera drift.

We used the eyetracking data to home in specifically on behavioral effects
related to \textit{covert} attention, as opposed to overt looking effects.
Specifically, we excluded from further analysis any images from trials where
participants shifted their gaze (for any non-zero amount of time) to any part
of the attended composite image during a presentation trial
(see Figs.~\gazeLocations,~and \excludedTrials).

@@

Our experiment comprised two testing conditions: a \textit{sustained} attention
condition and a \textit{variable} attention condition. Both experimental
conditions comprised a practice phase followed by a series of eight task
blocks. Each task block was in turn comprised of a presentation phase and a
memory phase. The practice and presentation phases differed across the two
experiments, and the memory phases were identical across the two experiments.
Both experiments were implemented using PsychoPy~\citep{PeirEtal19}.

@@

Several participants in pilot versions of our experiments reported that they
found it difficult to modulate the focus of their attention quickly on command.
We therefore designed a practice sequence to orient the participant to the
process of quickly modulating the focus of their attention without moving their
eyes. The experimenter remained in the testing room throughout the practice
phase and answered any questions about the experiment. The practice sequence
builds up incrementally to provide a gradual on-ramping for the participant
prior to beginning the main experimental tasks that we focused on in our
analyses.

@@

After completing the practice phase of the experiment, the participant read the
instructions for the task blocks (described next). The experimenter gave
participants a chance to ask any remaining questions about the experiment.
After answering the participant's questions, the experimenter calibrated the
eyetracker and exited the testing room.

@@

Participants viewed a succession of 10 attention cues (1.5~s), each followed by
a fixation cross (1~s), composite image (3~s), and a reaction time probe. The
attention cues were selected pseudorandomly across trials within each block,
with the constraints that no single attention cue pair could appear more than
three times across the 10 composite image pairs within a single task block.

@@

After the presentation phase of each task block, we asked the participant to
rate the familiarity (on a 1--4 scale, as during the practice phase) of a
succession of face and place images. Each image was preceded by a 1~s fixation
cross, and participants had up to 2~s to input their rating of each image.
Participants made a total of 40 familiarity judgements, about 20 face images
and 20 place images. Of these, 20 of the images (10 faces and 10 places) were
drawn randomly from the (attended and unattended) composite images that the
participant had viewed during the presentation phase. The remaining 20 images
(10 faces and 10 places) were novel images that the participant had not
encountered during any part of the experiment. At the end of each memory block,
the participant was given the opportunity to take a short break. When they were
ready to continue with the next task block, they indicated their readiness to
the experimenter. The experimenter then entered the testing room, re-ran the
eyetracker calibration sequence, and exited the testing room prior to the next
task block.

@@

We first wondered whether (and how) shifts in covert attention might affect
participants' ratings during the recognition memory task
(Fig.~\ref{fig:familiarity}). To ensure that our findings were not conflated
with where people were physically looking, we excluded from further analysis
any images presented during trials where the participant's gaze touched on any
part of the attended composite image (see \textit{Eyetraciking},
Figs.~\gazeLocations~and \excludedTrials). For the remaining trials, the
participants kept their gaze focused on a fixation cross at the center of the
screen while \textit{covertly} shifting the focus of their attention to the
cued category component of the composite image on the cued location. In other
words, during these remaining trials, participants' physical (external)
experiences of the face and place components of every presented composite image
remained relatively constant across trials (up to our ability to accurately
measure where participants were looking using the eyetracker).

Simply by encoding their prior experiences into memory, we reasoned that
participants should rate \textit{any} presented images as more familiar than
novel images, regardless of whether they were following the attention cues. We
confirmed that this prediction held in both the sustained ($t(29) = 8.856, p <
0.001$) and variable ($t(22) = 5.144, p < 0.001$) conditions. In addition, to
the extent that participants were following the attention cues, there
\textit{internal} experiences of each image should depend on their internal
focus of attention during each image presentation. For example, we expected
that the attended-category component of the composite image at the attended
location might be better recognized than the other composite image components.
Indeed, participants in both experimental conditions rated these ``fully
attended'' images as more familiar than category-matched image components from
unattended locations (sustained: $t(29) = 6.893, p < 0.001$; variable: $t(22) =
6.938, p < 0.001$), location-matched images from the unattended category
(sustained: $t(29) = 6.710, p < 0.001$; variable: $t(22) = 7.633, p < 0.001$),
unattended images that were neither from the attended category nor the attended
location (sustained: $t(29) = 8.470, p < 0.001$; variable: $t(22) = 7.256, p <
0.001$), or novel images they had never seen before (sustained: $t(29) =
10.259, p < 0.001$; variable: $t(22) = 7.874, p < 0.001$).

We also wondered whether the ways participants attended to or remembered the
images might depend on image-specific properties like the images' categories.
We repeated the analyis displayed in Figure~\ref{fig:familiarity} separately
for face and place images (Fig.~\ratingsByCategory). The same general patterns
held for each category, as when we combined across the two stimulus categories
as reported above. For example, fully attended face and place images were both
rated as more familiar than the category-matched images from the unattended
location (sustained: $t$s$(29) \geq 3.36, p$s$ \leq 0.002$; variable: $t$s$(22)
\geq 6.205, p$s$ \leq 0.001$), attended-location images from the unattended
category (sustained: $t$s$(29) \geq 5.886, p$s$ \leq 0.001$; variable:
$t$s$(29) \geq 4.277, p$s$ \leq 0.001$), images from the unattended category
and location (sustained: $t$s$(29) \geq 6.628, p$s$ \leq 0.001$; variable:
$t$s$(29) \geq 5.624, p$s$ \leq 0.001$), and novel images (sustained: $t$s$(29)
\geq 5.987, p$s$ \leq 0.001$; variable: $t$s$(29) \geq 5.132, p$s$ \leq
0.001$). Taken together, the above results suggest that what we remember is
guided in part by what we attended to, even after accounting for where we look
or what specifically we are looking at.

Splitting participants' responses to face versus place images also revealted
that participants often rated attended (and partially attended) place images as
more familiar than attention-matched face images (compare dark versus light
bars in Fig.~\ratingsByCategory). We hypothesized that this might be
explainable by some property of the relevant cognitive processes or by
properties of the stimuli themselves. To help elucidate this distinction, we
examined individual exemplars of the face and place images used in our paradigm
(Fig.~\ref{fig:stimuli}A). By design, the face images had consistent head
sizes, viewing angles, expressions, and so on. In contrast, the place images
varied more substantially across images. For example, some place images
depicted human-made structures; others depicted natural scenes; some depicted
indoor views; others depicted outdoor views; etc. This can also be seen by
averaging the pixel intensity values across images, separately for the face and
place stimuli (Fig.~\ref{fig:stimuli}B). Whereas the average face image retains
many of the landmarks characteristic of most faces (e.g., clearly defined hair,
eyes, nose, mouth, head shape, etc.), the average place image does not show
place-specific features as clearly, aside from a general tendancy for the tops
of place images to be lighter than the bottoms of place images. We also
computed the pairwise similarities across images from each stimulus category
(Fig.~\ref{fig:stimuli}C) and found that face images tended to be much more
similar to each other than place images (Fig.~\ref{fig:stimuli}D; $t(115258) =
254.764, p < 0.001$). This analysis indicated to us that our experimental
paradigm was not well-suited to identifying cognitively meaningful stimulus
category differences, since participants' category-specific judgements may be
confounded with within-category image similarity differences.

@@

  \caption{\textbf{Familiarity ratings by serial positions and attention
  level.} \textbf{A. Subsequent familiarity ratings by presentation position
  (sustained attention condition).} The curves' colors denote the attention
  levels of each presented image. The $x$-axis denotes the presentation
  positions of each image within the sequence of 10 composite image pairs
  during the run when it was presented. The $y$-axis denotes the average
  familiarity ratings later given to the corresponding items. \textbf{B.
  Subsequent familiarity ratings by presentation position (variable attention
  condition).} This panel is in the same format as Panel A, but displays
  ratings for the variable attention condition. \textbf{C. Familiarity ratings
  by memory probe position (sustained attention condition).} The curve's colors
  denote the attention levels (or novelty) of each probe image. The $x$-axis
  denotes the position of each probed image within the sequence of 40 images
  that participants judged during the memory phase of the experiemnt. The
  $y$-axis denotes the average familiarity ratings given to the corresponding
  probes. \textbf{D. Familiarity ratings by memory probe position (variable
  attention condition).} This panel is in the same format as Panel C, but
  displays ratings for the variable attention condition. All panels: error
  ribbons denote across-participant bootstrap-estimated 95\% confidence
  intervals.}

>>>>

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{hyperref}
\usepackage[sort]{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\usepackage{setspace,caption}
\doublespacing
\captionsetup[figure]{labelsep=period,labelfont=it,justification=justified,
 singlelinecheck=false,font=doublespacing}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\linenumbers

@@

\author{Kirsten Ziman\textsuperscript{$1, 2$},
Madeline R. Lee$^1$,
Alejandro R. Martinez$^1$,\\
Ethan D. Adner$^1$,
and
Jeremy R. Manning\textsuperscript{$1, *$}\\[0.1in]$^1$Dartmouth College\\
$^2$Princeton University\\
\textsuperscript{*}Address correspondence to jeremy.r.manning@dartmouth.edu}

@@

\noindent\textbf{Keywords: covert attention, volitional attention,
location-based attention, category-based attention, recognition memory}

@@

Our brains' cognitive systems detect and exploit patterns in our prior and
ongoing experiences, enabling us to function and adapt in an ever-changing
world. However we do not attend to or treat all types of remembered or incoming
information equally, and our ability to flexibly adapt our thinking and
behaviors can vary markedly with the specific set of concepts or tasks relevant
to a given setting or situation~\citep{ChunTurk07, AlyTurk17, HardNade09,
RangRitc12, JayaEtal23, HirsAly22, deBeEtal21, KeenEtal22, HakiEtal20,
AdamdeBe19}. There is also substantial variability across people with respect
to which aspects of experience (sensory, social, emotional, etc.) are noticed,
discriminated between, and acted upon~\citep{HuntEtal89}. This implies that the
same physical (objective) experience may give rise to very different perceived
(subjective) experiences across people~\citep{FreeSimo11, ChanEtal21a}.

The aspects of our experience we attend may be under our volitional control or
may be unconscious or automatic~\citep{JacoEtal92}. Both volitional and
unconscious attention may be expressed overtly, for example through intentional
eye movements~\citep{HoffSubr95} or covertly, without any volitional physical
change~\citep{EngbKlie03}. Prior work has explored the similarities and
differences in the neural basis of overt versus covert
attention~\citep{PosnEtal87, HuntKing03} as well as the behavioral and neural
underpinnings of volitional versus unconscious attention~\citep{DijkAart10} and
their differential effects on memory. There is a general consensus that
sustained volitional attention enhances memory relative to unconscious
attentional processes~\citep{UncaEtal11, TurkEtal13}. However, volitional
attention takes many forms, such as attention to particular spatial locations
or attention to particular visual features or other stimulus properties. How
different \textit{types} of volitional attention combine (or compete) to
enhance memory remains an open question. Volitional covert attention is of
particular interest in that it allows us to dynamically and intentionally
manipulate our experience, even when our sensory input remains largely
static~\citep[i.e., constant physical stimulus, retinal image, etc.;][]{YiEtal06,
OcraEtal99}.

Here we examine the ways two different types of volitional covert attention
interact to affect memory. We designed an experimental
paradigm~\citep[following][]{Posn80} that asked participants to attend to a
series of presented composite image pairs while keeping their gaze fixed on a
central point. The image pairs comprised a left and right image, each
constructed by blending an image of a face and place (indoor or outdoor scene).
The stimuli and presentation durations were held constant across the two
experimental conditions, but the conditions differed in how often we asked
participants to change the focus of their attention with respect to image
category (face versus place) and image location (left versus right). After the
participants attended to a series of images, we used a recognition memory test
to assess which aspects of the presented images had been encoded into memory.
In both conditions we found that the images participants covertly attended to
were better recognized than other images, supporting the notion that attention
enhances memory encoding~\citep[i.e., they rated attended images as more
familiar than unattended images;][]{Yone02}. To a lesser extent, participants
in both conditions also recognized partially attended images (e.g., images from
the attended category but unattended location, or images at the attended
location but from the unattended category). However, the ways the
category-based and location-based attention affected participants' memories for
these partially attended stimuli differed across the two experimental conditions.
These partial attention differences suggest that different forms of attention
affect memory encoding on different timescales.

@@

We ran a total of 53 participants in across two experimental conditions
(Fig.~\ref{fig:paradigm}). The two conditions differed in how often we
cued participants to change the focus of their attention. All code and
documentation pertaining to our experiment and analyses, along with the
experimental stimuli and data, may be downloaded from
\url{http://www.github.com/ContextLab/attention-memory-task}.

@@

  \caption{\textbf{Experimental paradigm.} \textbf{A.--D. Practice phase.}
  \textbf{A.} Composite face/place image. \textbf{B.} A single pair of
  composite images and a central fixation cross. \textbf{C.} One attention cue
  practice trial. \textbf{D.} Familiarity judgement practice trial. \textbf{E.
  Sustained attention condition.} Participants receive an attention cue,
  followed by a sequence of 10 composite image pairs (Panel E), and then they
  make of familiarity judgements on each of 40 face and place images, presented
  in sequence (Panel F). \textbf{G. Variable attention condition.} Participants
  study a succession of 10 composite image pairs, each preceded by an attention
  cue. After studying the images, they make a series of 40 familiarity
  judgements about face and place images, as in the sustained attention
  condition (Panel F). Note: illustrations are not drawn to scale; see the main
  text for sizing information.}

@@

A total of 53 Dartmouth College undergraduate students enrolled in our study
(30 in the sustained attention condition and 23 in the variable attention
condition). Following a pilot study using a similar experimental design, we
had aimed to enroll 30 participants in each condition. However, we fell short of
our enrollment target in the variable attention condition when in-person
testing was discontinued at our institution due to the COVID-19 pandemic. We
nonetheless chose to analyze and report on our findings from this
smaller-than-anticipated cohort of participants using all of the viable data we
were able to collect. All participants in our study had self-reported normal or
corrected-to-normal vision, memory, and attention. Participants gave written
consent to enroll in the study under a protocol approved by the Committee for
the Protection of Human Subjects at Dartmouth College.

We used a voluntary pre-experimental survey to collect self-reported
demographic information about each partipant. All 53 participants elected to
fill out the survey. Participants ranged in age from 18--21 years old (mean:
18.7 years; standard deviation: 0.8 years). Participants reported their genders
as female (34 participants), male (18 participants), and gender non-binary (1
participant). Participants reported their ethnicities as Not Hispanic or Latino
(44 participants), Hispanic or Latino (7 participants), or declined to respond
(2 participants). Participants reported their race as White (37 participants),
Asian (13 participants), American Indian or Alaska Native (4 participants),
Black or African American (2 participants), and Other (1 participant). Note
that each participant could report one or more racial categories, as they
deemed appropriate.

Forty-nine participants reported having no reading impairments, and 4
participants reported having reading impairments such as mild dyslexia. Fifty
participants reported having normal color vision and 3 reported having abnormal
color vision such as colorblindness. Fifty participants reported taking no
medications and having no recent injuries. One participant reported that they
had recently ``hit [their] head very hard.'' Another reported having taken
concerta (methylphenidate) in the past, but mentioned they had not taken it
recently. One participant reported using amphetamines regularly, but also
clarified that they had not used amphetamines on their testing day.

We also asked participants to self-report on their sleep, alertness, and coffee
consumption. Participants reported having gotten between 4 and 9 hours of sleep
on the night prior to testing (mean: 6.9 hours; standard deviation: 1.3 hours).
Participants reported their alertness at the time of testing, and we converted
their responses to point values as follows: ``Very alert'' (5 points), ``A
little alert'' (4 points), ``Neutral'' (3 points), ``A little sluggish'' (2
points), and ``Very sluggish'' (1 point). Across all participants, the full
range of alertness values were used (maximum: 5 points; minimum: 1 point; mean:
3.4 points; standard deviation: 1.0 point). Participants reported having
consumed between 0 and 2 cups of coffee so far on their testing day (mean: 0.3
cups; standard deviation: 0.5 cups).

@@

Participants viewed photographs of faces, places, and composite images each
comprising an equal blend of one face image and one place image. The pool of
360 face images included photographs of adult human male and female faces
selected from the FERET database~\citep{PhilEtal98}. The pool of 360 place
images included photographs of indoor and outdoor places selected from the SUN
database~\citep{XiaoEtal10}. The images we used from both databases came from a
stimulus subset that was manually curated by Megan deBettencourt (personal
communication). All images were resized to $256 \times 256$ pixels, converted
to greyscale, and processed so that every image was matched for mean contrast
and intensity. We selected 20 face images and 20 place images from the stimulus
pool to use in the instructional and practice phases of the experiment
(Fig.~\ref{fig:paradigm}A--D).

@@

 Our experiment was conducted in a sound- and light-attenuated testing room
 containing a chair, desk, and 27-inch iMac desktop computer (display
 resolution: $2048 \times 1152$). The participant sat in the chair and rested
 their chin on a chin rest located 60~cm from the display. The active portion
 of the display screen occupied 52.96$^\circ$ (width) and 31.28$^\circ$
 (height) of the participant's field of view from the chin rest. Stimuli were
 sized to occupy $6.7^\circ$ (width and height) of the participant's field of
 view from the chin rest. We maintained a black background (with any text or
 cues displayed in white) throughout the experiment.

\subsection*{Eye-tracking}

We recorded participants' eye gaze positions using a desk-mounted video-based
eye tracker with a spatial resolution of 0.1$^\circ$ visual angle root mean
squared error and a sampling rate of 30~Hz (Eye Tribe, The Eye Tribe,
Copenhagen, Denmark). We calibrated the eye tracker using a 9-point gaze
pattern. As described below, we re-calibrated the eye tracker at regular
intervals throughout the experiment to protect against camera drift.

We used the eye-tracking data to home in specifically on behavioral effects
related to \textit{covert} attention, as opposed to overt looking effects.
Specifically, we excluded from further analysis any images from trials where
participants shifted their gaze (for any non-zero amount of time) to any part
of the attended composite image during a presentation trial
(see Figs.~\gazeLocations,~and \excludedTrials).

@@

Our experiment comprised two testing conditions: a \textit{sustained} attention
condition and a \textit{variable} attention condition. Both experimental
conditions comprised a practice phase followed by a series of eight task
blocks. Each task block was in turn comprised of a presentation phase and a
memory phase. The practice and presentation phases differed across the two
conditions, and the memory phases were identical across the two conditions.
We implemented (coded) the experiment using PsychoPy~\citep{PeirEtal19}.

@@

Several participants in pilot versions of our experiment reported that they
found it difficult to modulate the focus of their attention quickly on command
without moving their eyes. We therefore designed a practice sequence to orient
the participant to the process of quickly modulating their focus of covert
attention. The experimenter remained in the testing room throughout the
practice phase and answered any questions about the experiment. The practice
sequence builds up incrementally to provide a gradual on-ramping for the
participant prior to beginning the main experimental tasks that we focused on
in our analyses.

@@

After completing the practice phase of the experiment, the participant read the
instructions for the task blocks (described next). The experimenter gave
participants a chance to ask any remaining questions about the experiment.
After answering the participant's questions, the experimenter calibrated the
eye tracker and exited the testing room.

@@

Participants viewed a succession of 10 attention cues (1.5~s), each followed by
a fixation cross (1~s), composite image (3~s), and a reaction time probe. The
attention cues were selected randomly across trials within each block.

@@

After the presentation phase of each task block, we asked the participant to
rate the familiarity (on a 1--4 scale, as during the practice phase) of a
succession of face and place images. Each image was preceded by a 1~s fixation
cross, and participants had up to 2~s to input their rating of each image.
Participants made a total of 40 familiarity judgements, about 20 face images
and 20 place images. Of these, 20 of the images (10 faces and 10 places) were
drawn randomly from the (attended and unattended) composite images that the
participant had viewed during the presentation phase. The remaining 20 images
(10 faces and 10 places) were novel images that the participant had not
encountered during any part of the experiment. At the end of each memory phase,
the participant was given the opportunity to take a short break. When they were
ready to continue with the next task block, they indicated their readiness to
the experimenter. The experimenter then entered the testing room, re-ran the
eye tracker calibration sequence, and exited the testing room prior to the next
task block.

@@

We first wondered whether (and how) shifts in covert attention might affect
participants' ratings during the recognition memory task
(Fig.~\ref{fig:familiarity}). To ensure that our findings were not conflated
with where people were physically looking, we excluded from further analysis
any images presented during trials where the participant's gaze touched on any
part of the attended composite image (see \textit{Eye-tracking},
Figs.~\gazeLocations~and \excludedTrials). For the remaining trials, the
participants kept their gaze focused on a fixation cross at the center of the
screen while \textit{covertly} shifting the focus of their attention to the
cued category component of the composite image at the cued location. In other
words, during these remaining trials, participants' physical (external)
experiences of the face and place components of every presented composite image
remained relatively constant across trials (up to our ability to accurately
measure where participants were looking using the eye tracker).

Simply by encoding their prior experiences into memory, we reasoned that
participants should rate \textit{any} presented images as more familiar than
novel images, regardless of whether they were following the attention cues. We
confirmed that this prediction held in both the sustained ($t(29) = 8.856, p <
0.001$) and variable ($t(22) = 5.144, p < 0.001$) conditions. In addition, to
the extent that participants were following the attention cues, their
\textit{internal} experiences of each image should depend on their internal
focus of attention during each image presentation. For example, we expected
that the attended-category component of the composite image at the attended
location might be better recognized than the other composite image components.
Indeed, participants in both experimental conditions rated these ``fully
attended'' images as more familiar than category-matched image components from
unattended locations (sustained: $t(29) = 6.893, p < 0.001$; variable: $t(22) =
6.938, p < 0.001$), location-matched images from the unattended category
(sustained: $t(29) = 6.710, p < 0.001$; variable: $t(22) = 7.633, p < 0.001$),
unattended images that were neither from the attended category nor the attended
location (sustained: $t(29) = 8.470, p < 0.001$; variable: $t(22) = 7.256, p <
0.001$), or novel images they had never seen before (sustained: $t(29) =
10.259, p < 0.001$; variable: $t(22) = 7.874, p < 0.001$).

We also wondered whether the ways participants attended to or remembered the
images might depend on image-specific properties like the images' categories.
We repeated the analysis displayed in Figure~\ref{fig:familiarity} separately
for face and place images (Fig.~\ratingsByCategory). The same general patterns
reported above also held for each stimulus category individually. For example,
fully attended face and place images were both rated as more familiar than the
category-matched images from the unattended location (sustained: $t$s$(29) \geq
3.366, p$s$ \leq 0.002$; variable: $t$s$(22) \geq 4.773, p$s$ \leq 0.001$),
attended-location images from the unattended category (sustained: $t$s$(29)
\geq 5.886, p$s$ \leq 0.001$; variable: $t$s$(29) \geq 4.277, p$s$ \leq
0.001$), images from the unattended category and location (sustained: $t$s$(29)
\geq 6.628, p$s$ \leq 0.001$; variable: $t$s$(29) \geq 5.624, p$s$ \leq
0.001$), and novel images (sustained: $t$s$(29) \geq 5.987, p$s$ \leq 0.001$;
variable: $t$s$(29) \geq 5.132, p$s$ \leq 0.001$). Taken together, the above
results suggest that what we remember is guided in part by what we attended to,
even after accounting for where we look or what specifically we are looking at.

Splitting participants' responses to face versus place images also revealed
that participants often rated attended (and partially attended) place images as
more familiar than attention-matched face images (compare dark versus light
bars in Fig.~\ratingsByCategory). We hypothesized that this might be
explainable by some property of the relevant cognitive processes or by
properties of the stimuli themselves. To help elucidate this distinction, we
examined individual exemplars of the face and place images used in our paradigm
(Fig.~\ref{fig:stimuli}A). By design, the face images had consistent head
sizes, viewing angles, expressions, and so on. In contrast, the place images
varied more substantially across images. For example, some place images
depicted human-made structures; others depicted natural scenes; some depicted
indoor views; others depicted outdoor views; etc. This can also be seen by
averaging the pixel intensity values across images, separately for the face and
place stimuli (Fig.~\ref{fig:stimuli}B). Whereas the average face image retains
many of the landmarks characteristic of most faces (e.g., clearly defined hair,
eyes, nose, mouth, head shape, etc.), the average place image does not show
place-specific features as clearly, aside from a general tendency for the tops
of place images to be lighter than the bottoms of place images. We also
computed the pairwise similarities across images from each stimulus category
(Fig.~\ref{fig:stimuli}C) and found that face images tended to be much more
similar to each other than place images (Fig.~\ref{fig:stimuli}D; $t(115258) =
254.764, p < 0.001$). This analysis indicated to us that our experimental
paradigm was not well-suited to identifying cognitively meaningful stimulus
category differences, since participants' category-specific judgements may be
confounded with within-category image similarity differences.

@@

  \caption{\textbf{Familiarity ratings by serial positions and attention
  level.} \textbf{A. Subsequent familiarity ratings by presentation position
  (sustained attention condition).} The curves' colors denote the attention
  levels of each presented image. The $x$-axis denotes the presentation
  positions of each image within the sequence of 10 composite image pairs
  during the run when it was presented. The $y$-axis denotes the average
  familiarity ratings later given to the corresponding items. \textbf{B.
  Subsequent familiarity ratings by presentation position (variable attention
  condition).} This panel is in the same format as Panel A, but displays
  ratings for the variable attention condition. \textbf{C. Familiarity ratings
  by memory probe position (sustained attention condition).} The curve's colors
  denote the attention levels (or novelty) of each probe image. The $x$-axis
  denotes the position of each probed image within the sequence of 40 images
  that participants judged during the memory phase of the experiment. The
  $y$-axis denotes the average familiarity ratings given to the corresponding
  probes. \textbf{D. Familiarity ratings by memory probe position (variable
  attention condition).} This panel is in the same format as Panel C, but
  displays ratings for the variable attention condition. All panels: error
  ribbons denote across-participant bootstrap-estimated 95\% confidence
  intervals.}
